---
title: 'EDS230/ESM232: Modeling Growth'
author: "Juliet Cohen, Dani Sclafani, Nikole Vannest"
date: '2022-05-13'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(sensitivity)
library(tidyverse)
library(deSolve)
```

### Modeling Growth

Consider the following model of forest growth (where forest size in measured in units of carbon (`C`))

`dC/dt  = r*C`  for forests where `C` is below a threshold canopy closure
`dC/dt = g*(1- C/K)` for forests where carbon is at or above the threshold canopy closure and `K` is a carrying capacity in units of carbon

- The size of the forest (`C`), Canopy closure threshold and carrying capacity are all in units of `kg carbon` 
- You could think of the canopy closure threshold as the size of the forest at which growth rates change from exponential to linear.
- You can think of `r` as early exponential growth rate and `g` as the linear growth rate once canopy closure has been reached.

1. Implement this model in R (as a **differential equation**) (Grading 25% - we will look for correct implementation but also good programming style - meaningful variable names, comments/documentation throughout)

```{r}
# r = early exponential growth rate before canopy closure, default = 0.01
# g = linear growth rate (once canopy closure has been reached), default = 2, units = kg/year
# C = size of forest, default = 10, units = kgC which is kilograms of carbon
# K = carrying capacity, default = 250
# we do not need to define time in the function itself, that is only necessary for inserting the function in to ODE solver and doing sobol sensitivity

forest_carbon <- function(r, C_initial, g) {
   carbon = r*C_initial # example of plot here: https://naomitague.github.io/ESM232_course/lectures/lecture9_simpledynamic.html#(3)
   carbon = g*(1 - C_initial/K)
   return(list(carbon)) # the ODE solver needs the output to be a list
}

# maybe we need to use an if-else statement?
forest_carbon_if_else <- function(r, C_initial, g) {
  if (C_initial < canopy_closure_threshold){
  carbon = r*C_initial
  }
  else {
  carbon = g * (1 - C_initial/K)
  }
  return(list(carbon)) # the ODE solver needs the output to be a list
}
```

```{r}
# test the if-else function on a value below threshold (the given C_initial)
canopy_closure_threshold <- 50

forest_carbon_if_else(r = 0.01, C_initial = 10, g = 2)
# carbon = 0.1, a static value because we are not running function over time but rather at the first time step

# test the if-else function on a value above the threshold (above the given C_initial)
K = 250

forest_carbon_if_else(r = 0.01, C_initial = 60, g = 2)
# carbon = 1.52, a static value because we are not running function over time but rather at the first time step
```

```{r}
# read in Namoi's functions to first test code on a function that we know works
# changed each P to a C for carbon

# dexppop <- function (time, C, r) {
#     dexpop = r * C
#     return(list(dexpop)) # the ODE solver needs the output to be a list
# }
# 
# dexppopK <- function (time, C, parms) {
#   dexpop = parms$r * C * (1 - C/parms$K)
#   return(list(dexpop)) # the ODE solver needs the output to be a list
# }
```

2. Run the model for 300 years (using the ODE solver) starting with an initial forest size of 10 kgC, and using the following parameters:

canopy closure threshold of 50 kgC 

K = 250 kgC (carrying capacity) 

r =  0.01 (exponential growth rate before before canopy closure)

g = 2 kg/year (linear growth rate after canopy closure)

Graph the results. (15% - we look for an appropriate graphs and good visualization practice - labels, etc)

```{r}
# code refenced from the lecture slides here: https://naomitague.github.io/ESM232_course/lectures/lecture10_moredynamic.html#(7)
# define the initial forest size in units of kilograms of carbon
C_initial = 10

# define time range in years
simtimes = seq(from = 1, to = 300, by = 5)

# for now the parameter(s) can be static values, but when we feed these paramaters into sobol sensitivity analysis later we will need to make them ranges with means and standard deviation
r = 0.01
K = 250
g = 2
static_parms <- c(r) # naomi did this in her code (https://naomitague.github.io/ESM232_course/lectures/lecture10_moredynamic.html#(7)), I am not sure how we should change this variable for our purposes

# run the ODE solver
C_over_time = ode(y = C_initial, 
            times = simtimes, 
            func = forest_carbon_if_else,
            parms = static_parms)

# rename columns for clarity: change "time" to "Year" and "1" to "Carbon"
colnames(C_over_time) = c("Year", "Carbon")

head(C_over_time)
```

```{r}
# code refenced from the lecture slides here: https://naomitague.github.io/ESM232_course/lectures/lecture10_moredynamic.html#(7)
# define the initial forest size in units of kilograms of carbon
C_initial = 10

# define time range in years
simtimes = seq(from = 1, to = 300, by = 5)

# for now the parameter(s) can be static values, but when we feed these paramaters into sobol sensitivity analysis later we will need to make them ranges with means and standard deviation
r = 0.01
K = 250
g = 2
static_parms <- c(r, K, g) # naomi did this in her code, but she only had c(r) (https://naomitague.github.io/ESM232_course/lectures/lecture10_moredynamic.html#(7)) and here I added K and g to the vector of values, but the output is the same either way, I am not sure how we should change this variable for our purposes
# run the ODE solver
C_over_time = ode(y = C_initial, 
            times = simtimes, 
            func = forest_carbon_if_else,
            parms = static_parms)

# rename columns for clarity: change "time" to "Year" and "1" to "Carbon"
colnames(C_over_time) = c("Year", "Carbon")

head(C_over_time)

ggplot(as.data.frame(C_over_time), aes(Year, Carbon)) + 
  geom_point() + 
  labs(y = "Carbon",
       x = "Years")
# this graph makes no sense ecologically, something must be wrong in the function itself 
```


3. Run a Sobol sensitivity analysis that explores how the estimated **maximum and mean forest size** (e.g maximum and mean values of `C` over the 300 years) varies with the pre canopy closure growth rate (`r`) and post-canopy closure growth rate (`g`) and canopy closure threshold and carrying capacity (`K`)

Assume that parameters are all normally distributed with means as given above and standard deviation of 10% of mean value

Graph the results of the sensitivity analysis as a box plot of maximum forest size and a plot of the two Sobol indices (S and T). (25% - correct implementation of Sobol and graph We also look for good graphing style -labels etc)

```{r}
# want to learn about sensitivity to pre-canopy growth rate (r), post-canopy closure growth rate (g), and canopy closure threshold and carrying capacity (K)

## CREATE FIRST SET OF SAMPLES
# define the number of parameters (called "observations" in the documentation for the rnorm() function)
np = 300 # i THINK this should be 300 because that is the number of time steps we were given, not sure if it would run otherwise 
# define K's standard deviation, which is 10% of the mean of 250
K_sd <-  0.1*250
# define K, using the default mean provided, the calculated standard deviation, and the number of observations  
K <-  rnorm(mean = 250, sd = K_sd, n = np)
# define r's standard deviation, which is 10% of the mean of 0.01
r_sd <- 0.1*0.01
# define r, pre-canopy closure growth rate, using the default mean provided, the calculated standard deviation, and the number of observations  
r = rnorm(mean = 0.01, sd = r_sd, n = np)
# define g's standard deviation, which is 10% of the mean of 2
g_sd <- 0.1*2
# define g, the post-canopy closure growth rate, using the default mean provided, the calculated standard deviation, and the number of observations
g = rnorm(mean = 2, sd = g_sd, n = np)
# bind all 3 parameters together
X1 = cbind.data.frame(r = r, K = K, g = g)

## CREATE SECOND SET OF SAMPLES (for sensitivity analysis)
# these are the same arguments that defined K and r for the first set sample set
K <-  rnorm(mean = 250, sd = K_sd, n = np)
r = rnorm(mean = 0.01, sd = r_sd, n = np)
g = rnorm(mean = 2, sd = g_sd, n = np)
X2 = cbind.data.frame(r = r, K = K, g = g)

# run sensitivity analysis
# here I chose 600 for the number of bootstrap replicates because in Naomi's example, she used np = 100 and nboot = 300, so I figure that nboot needs to be higher than np which we have as 300
sens_C = sobolSalt(model = NULL, X1, X2, nboot = 600)

# lets add names to the columns
colnames(sens_C$X) = c("r","K", "g")

# check that the columns and names look correct
head(sens_C$X)

# note: `simtimes` and `C_initial` objects were already defined earlier in this Rmd around line 125 when we tested the `forest_carbon_if_else()` function

p_wrapper = function(r, K, C_initial, simtimes, func, metrics_func) {
    parms = list(r = r, K = K)
    result = ode(y = C_initial, times = simtimes, func = func, parms = parms) 
    colnames(result) = c("time", "C")
  # get metrics
    metrics = as.data.frame(result) %>% metrics_func()
  return(metrics)
}

# this code does not run until we get the function to compute metrics...here naomi had compute_pop_metrics for the metrics_func argument but I could not find that anywhere so I am not sure what to put there for our purposes
allresults = as.data.frame(sens_C$X) %>%
  pmap(p_wrapper,
       C_initial = C_initial,
       simtimes = simtimes,
       func = dexppopK,
       metrics_func = compute_pop_metrics)
```

## THIS IS WHERE JULIET STOPPED CHANGING NAOMI'S CODE TO ACCOMODATE OUR ASSIGNMENT

```{r}
# extract out results from pmap into a data frame
allres = allresults %>% 
  map_dfr(`[`,c("max_carbon", "dyear"))

# create boxplots
tmp = allres %>% 
  gather(key="metric", value="value")

ggplot(tmp, aes(metric, value, col = metric)) +
  geom_boxplot()
```

```{r}
# compute sobol indicies (separately for the two different outputs)
sens_C_maxpop = sensitivity::tell(sens_C, allres$maxpop)

# first-order indices (main effect without co-variance)
sens_C_maxpop$S
```


```{r}
# total sensitivity index -note that this partitions the output variance - so values sum to 1
sens_C_maxpop$T
```


In 2-3 sentences, discuss what the results of your simulation might mean for climate change impacts on forest growth (e.g think about what parameters climate change might influence ). (25% - we look for reasonable discussion that uses the results from your analysis and give extra points for discussions that offer particularly creative or insightful commentary)

Submit R markdown with model implementation, graphs and sensitivity analysis and R file with your model (Final 10% for well-organized, clearly documented R markdown)

