---
title: 'EDS230/ESM232: Modeling Growth'
author: "Juliet Cohen, Dani Sclafani, Nikole Vannest"
date: '2022-05-13'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(sensitivity)
library(tidyverse)
library(deSolve)
```

### Modeling Growth

Consider the following model of forest growth (where forest size in measured in units of carbon (`C`))

`dC/dt  = r*C`  for forests where `C` is below a threshold canopy closure
`dC/dt = g*(1- C/K)` for forests where carbon is at or above the threshold canopy closure and `K` is a carrying capacity in units of carbon

- The size of the forest (`C`), Canopy closure threshold and carrying capacity are all in units of `kg carbon` 
- You could think of the canopy closure threshold as the size of the forest at which growth rates change from exponential to linear.
- You can think of `r` as early exponential growth rate and `g` as the linear growth rate once canopy closure has been reached.

1. Implement this model in R (as a **differential equation**) (Grading 25% - we will look for correct implementation but also good programming style - meaningful variable names, comments/documentation throughout)

```{r}
# r = early exponential growth rate before canopy closure, default = 0.01
# g = linear growth rate (once canopy closure has been reached), default = 2, units = kg/year
# C = size of forest, default = 10, units = kgC which is kilograms of carbon
# K = carrying capacity, default = 250

forest_growth <- function(time, r, C_initial, g = 2) {
   growth = r*C_initial # example of plot here: https://naomitague.github.io/ESM232_course/lectures/lecture9_simpledynamic.html#(3)
   growth = g*(1 - C_initial/K)
   return(list(growth))
}

# maybe we need to use an if-else statement?
# define threshold_canopy_colsure, i have no idea what that would be, this value is a stab in the dark
threshold_canopy_colsure <- 60
forest_growth_if_else <- function(time, r, C_initial, g = 2) {
  if (C_initial < threshold_canopy_closure){
  growth = r*C_initial
  }
  else {
  growth = g * (1 - C_initial/forest_params$K)
  }
  return(list(growth))
}
```

```{r}
# read in her function to first test code on a function that we know works...?
dexppopK <- function (time, C, parms) {
  dexpop = parms$r * C * (1 - C/parms$K)
  return(list(dexpop))
}
```

2. Run the model for 300 years (using the ODE solver) starting with an initial forest size of 10 kgC, and using the following parameters:

canopy closure threshold of 50 kgC 

K = 250 kgC (carrying capacity) 

r=  0.01 (exponential growth rate before before canopy closure)

g = 2 kg/year (linear growth rate after canopy closure)

Graph the results. (15% - we look for an appropriate graphs and good visualization practice - labels, etc)

```{r}
# define the initial forest size in units of kilograms of carbon
C_initial = 10

# define time range in years
simtimes = seq(from = 1, to = 300, by = 5)
```

3. Run a Sobol sensitivity analysis that explores how the estimated maximum and mean forest size (e.g maximum and mean values of C over the 300 years) varies with the pre canopy closure growth rate (r) and post-canopy closure growth rate (g) and canopy closure threshold and carrying capacity(K)

Assume that parameters are all normally distributed with means as given above and standard deviation of 10% of mean value

Graph the results of the sensitivity analysis as a box plot of maximum forest size and a plot of the two Sobol indices (S and T). (25% - correct implementation of Sobol and graph We also look for good graphing style -labels etc)

```{r}
# want to learn about sensitivity to pre-canopy growth rate (r), post-canopy closure growth rate (g), and canopy closure threshold and carrying capacity(K)

## CREATE FIRST SET OF SAMPLES
# define the number of parameters, i dont know what this does in the rnorm() function
np = 300
# define K's standard deviation, which is 10% of the mean of 250
K_sd <-  0.1*250
# define K, using the default mean provided, the calculated standard deviation, and the number of time steps 
K <-  rnorm(mean = 250, sd = K_sd, n = np)
# define r's standard deviation, which is 10% of the mean of 0.01
r_sd <- 0.1*0.01
# define r, using the default mean provided, the calculated standard deviation, and the number of time steps 
r = rnorm(mean = 0.01, sd = r_sd, n = np)
X1 = cbind.data.frame(r = r, K = K)

## CREATE SECOND SET OF SAMPLES (for sensitivity analysis)
# these are the same arguments that defined K and r for the first set sample set
K <-  rnorm(mean = 250, sd = K_sd, n = np)
r = rnorm(mean = 0.01, sd = r_sd, n = np)
X2 = cbind.data.frame(r = r, K = K)

# run sensitivity analysis
# here I chose 600 for the number of bootstrap replicates because in Naomi's example, she used np = 100 and nboot = 300, so I figure that nboot needs to be higher than np
sens_C = sobolSalt(model = NULL, X1, X2, nboot = 600)

# lets add names to the columns
colnames(sens_C$X) = c("r","K")

# check that the columns and names look correct
head(sens_C$X)

# note: `simtimes` and `C_initial` objects were already defined earlier in this Rmd

p_wrapper = function(r, K, C_initial, simtimes, func, metrics_func) {
    parms = list(r = r, K = K)
    result = ode(y = C_initial, times = simtimes, func = func, parms = parms) 
    colnames(result) = c("time", "C")
  # get metrics
    metrics = as.data.frame(result) %>% metrics_func()
  return(metrics)
}

# this code does not run until we get the function compute_pop_metrics
# i am not sure if we what function we will use to replace comput_pop_metrics for this assignment
allresults = as.data.frame(sens_C$X) %>%
  pmap(p_wrapper,
       C_initial = C_initial,
       simtimes = simtimes,
       func = dexppopK,
       metrics_func = compute_pop_metrics)

# extract out results from pmap into a data frame
allres = allresults %>% 
  map_dfr(`[`,c("maxpop", "dyear"))

# create boxplots
tmp = allres %>% 
  gather(key="metric", value="value")

ggplot(tmp, aes(metric, value, col = metric)) +
  geom_boxplot()
```

```{r}
# compute sobol indicies (separately for the two different outputs)
sens_C_maxpop = sensitivity::tell(sens_C, allres$maxpop)

# first-order indices (main effect without co-variance)
sens_C_maxpop$S
```


```{r}
# total sensitivity index -note that this partitions the output variance - so values sum to 1
sens_C_maxpop$T
```


In 2-3 sentences, discuss what the results of your simulation might mean for climate change impacts on forest growth (e.g think about what parameters climate change might influence ). (25% - we look for reasonable discussion that uses the results from your analysis and give extra points for discussions that offer particularly creative or insightful commentary)

Submit R markdown with model implementation, graphs and sensitivity analysis and R file with your model (Final 10% for well-organized, clearly documented R markdown)

